---
title: "Algebra in R"
output:
  html_notebook: default
  html_document: default
---

```{r setup, include=FALSE, error=TRUE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA, error = TRUE)
```


https://www.math.uh.edu/~jmorgan/Math6397/day13/LinearAlgebraR-Handout.pdf


All vectors are column vectors unless specified.

```{r}
a <- matrix(c(1, 3, 2), ncol=1)
a
```

a = $\begin{pmatrix}
 2  \\ 
 1  \\
 3
 \end{pmatrix}$
 
 
### Transpose of vectors

```{r}
a
```


Transpose of `a`
```{r}
t(a)
```

### Sum of vectors

```{r}
b <- c(2, 8, 9)
```

```{r}
a + b
```

> Only vectors of the same dimension can be added



### Inner product of vectors
It is a number not a vector.
```{r}
sum(a * b)
```

### Length of a vector (norm)

$\left \| a \right \| = \sqrt {a.a} = \sqrt \sum_{i=1}^{n} a_i^2$

```{r}
sqrt(sum(a * a))
```

### The -vector and 1-vector

```{r}
rep(0, 5)
```

```{r}
rep(1, 5)
```

### Orthogonal (perpedicular) vectors
Two vectors are perpendicular is the inner product is zero.
```{r}

v1 <- c(1, 1)
v2 <- c(-1, 1)

sum(v1 * v2)
```

## Matrices

### Matrices
```{r}
A <- matrix(c(1, 3, 2, 2, 8, 9), ncol = 3)
A
```

Note that the numbers 1, 3, 2, 2, 8, 9 are read into the matrix column–by–
column. To get the numbers read in row–by–row do

```{r}
A2 <- matrix(c(1, 3, 2, 2, 8, 9), ncol = 3, byrow = T)
A2
```

### Multiplying a matrix by a number

```{r}
A <- matrix(c(1, 3, 2, 2, 8, 9), ncol = 2)
7 * A
```

### Transpose of a matrix
```{r}
A
```
```{r}
t(A)
```

### Sum of matrices

```{r}
A <- matrix(c(1, 3, 2, 2, 8, 9), ncol = 2)
B <- matrix(c(5, 8, 3, 4, 2, 7), ncol = 2)

A+B
```

### Multiplication of a matrix and a vector

```{r}
a <- c(5, 8)

A %*% a
```

Note the difference to:

```{r}
A * a
```

### Multiplication of matrices

```{r}
C <- matrix(c(5, 4, 8, 2), ncol =2, byrow = T)
C
```

```{r}
A %*% C
```

> Note that the product AB can only be formed if the number of rows in B and
the number of columns in A are the same. In that case, A and B are said to
be conforme.
In general AB and BA are not identical.


##n Some special matrices
- An n × n matrix is a square matrix

– A matrix A is symmetric if A = A>.

– A matrix with 0 on all entries is the 0–matrix and is often written simply
as 0.

– A matrix consisting of 1s in all entries is of written J.

– A square matrix with 0 on all off–diagonal entries and elements d1, d2, . . . , dn
on the diagonal a diagonal matrix and is often written diag{d1, d2, . . . , dn}

– A diagonal matrix with 1s on the diagonal is called the identity matrix
and is denoted I. The identity matrix satisfies that IA = AI = A.

### 0-matrix
```{r}
matrix(0, nrow = 2, ncol = 3)
```

### 1-matrix
```{r}
matrix(1, nrow = 2, ncol = 3)
```

### Diagonal matrix
```{r}
diag(c(1, 2, 3))
```

### Identity matrix
```{r}
diag(1, 3)
```

### Applying `diag` to a matrix

```{r}
diag(diag(c(1,2,3)))
```

```{r}
diag(A)
```

## Inverse of matrices

```{r}
A <- matrix(c(1, 2, 3,4), ncol =2)
B <- matrix(c(-2, 1, 1.5, -0.5), ncol =2)
A
B
```
 
The identity matrix should be the result of A %*% B :
```{r}
A %*% B
```

### Find the inverse of a matrix

```{r}
B <- solve(A)
B
```

### Solving systems of linear erquations


For these two equations:

$x_1 + 3x_2 = 7 \\
2x_1 + 4x_2 = 10$

They can be written as:

$\begin{pmatrix}
 1 & 3 \\
 2 & 4
\end{pmatrix}
\begin{pmatrix}
  x_{1} \\
  x_{2} 
 \end{pmatrix}=
\begin{pmatrix}
  7 \\
  10  
\end{pmatrix}$

or:

$\mathbf{A}x = b$


Which we can express as:

$x = A^{-1} b$

since we know that:

$A^{-1} A = I$ and $Ix = x$

then:

$A^{-1} \mathbf{A} x = x$

which finally yields:

$A^{-1} b = x$

### Example of equation:

```{r}
A <- matrix(c(1, 2, 3, 4), ncol=2)
b <- matrix(c(7, 10), ncol=1)
```

```{r}
Ai <- solve(A)
Ai %*% b
```

The solution then is:

$\begin{pmatrix}
  x_{1} \\
  x_{2} 
 \end{pmatrix} = 
\begin{pmatrix}
  1 \\
  2
 \end{pmatrix}$
 
 
 
 

