---
title: "RcppEigen Examples"
output:
  html_document: default
  html_notebook: default
  pdf_document: default
---

```{r setup, include=FALSE, error=TRUE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA, error = TRUE)
```


## Show the current Version

```{r}
library(RcppEigen)

pkgVersion <- packageDescription("RcppEigen")$Version
pkgDate <- packageDescription("RcppEigen")$Date

pkgVersion
pkgDate
```


```{r}

RcppSamp <- function(X) {
  stopifnot(is.numeric(X <- as.matrix(X)),
            (nc <- ncol(X)) > 1L,
            all(X >= 0))
  .Call(CppSamp, X)
}

```


# Max over columns
source: http://stackoverflow.com/questions/27490659/rcppeigen-going-from-inline-to-a-cpp-function-in-a-package-and-map



## Using RcppEigen to find the max over columns

#### Long version
Assigns code to R variable `maxOverColCpp` Uses mixed C++.
```{r message=FALSE}
library(Rcpp); 
library(inline); 
library(RcppEigen);

maxOverColCpp <- '
    using Eigen::Map;
    using Eigen::MatrixXd;

   // Map the double matrix AA from R
   const Map<MatrixXd> A(as<Map<MatrixXd> >(AA));

   // evaluate and columnwise maximum entry of A
   const MatrixXd Amax(A.colwise().maxCoeff());
   return wrap(Amax);
'

```

Assign the inline C++ function:
```{r}
rcppeigen_max_over_columns <- cxxfunction(signature(AA = "matrix"), maxOverColCpp, plugin = "RcppEigen")

```

Create a matrix for the example:
```{r}
M <- matrix(c(1,2,3,4,5,6,7,8,9), ncol=3)
M
```

Run the inline C++ function:
```{r}
rcppeigen_max_over_columns(M)
```


## Max over columns: short version.
Uses C++ code.
This can even be saved as a C++ file (.cpp). Add `engine='Rcpp'` if want to run it in a notebook chunk.

```{r engine='Rcpp'}
// we only include RcppEigen.h which pulls Rcpp.h in for us
#include <RcppEigen.h>

using namespace Eigen;

// via the depends attribute we tell Rcpp to create hooks for
// RcppEigen so that the build process will know what to do
//
// [[Rcpp::depends(RcppEigen)]]
// [[Rcpp::export]]
MatrixXd  rcppeigen_max_over_columns3(const MatrixXd  & A) {
    MatrixXd Amax = A.colwise().maxCoeff();
    return Amax;
}
```

Run the C++ function:
```{r}
library(RcppEigen)

rcppeigen_max_over_columns3(M)
```

# Transpose
Uses R code assigned to variable `transCpp` and a call to `cxxfunction`.
the transpose() method for the `Eigen::MatrixXi` class is used to return the transpose of the supplied matrix. The R matrix in the SEXP AA is first mapped to an Eigen::MatrixXi object, and then the matrix At is constructed from its transpose and returned to R.

```{r}
library(Rcpp); 
library(inline); 
library(RcppEigen);

transCpp <- '
    using Eigen::Map;
    using Eigen::MatrixXi;
    
// Map the integer matrix AA from R
    const Map<MatrixXi> A (as<Map<MatrixXi> >(AA));
    
// evaluate and return the transpose of A
    const MatrixXi
    At(A.transpose());
    return wrap(At);
'
```

The R snippet below compiles and links the C++ code segment. The actual work is done by the function `cxxfunction` from the inline package which compiles, links and loads code written in C++ and supplied as a character variable. This frees the user from having to know about compiler and linker details and options, which makes “exploratory programming” much easier. Here the program piece to be compiled is stored as a character variable named `transCpp`, and `cxxfunction` creates an executable function which is assigned to `ftrans`. 

This new function is used on the matrix $A$ shown above, and one can check that it works as intended by comparing the output to an explicit transpose of the matrix argument.

```{r}
ftrans <- cxxfunction(signature(AA = "matrix"), transCpp, plugin = "RcppEigen")
```


```{r}
(A <- matrix(1:6, ncol = 2))
```

```{r}
str(A)
```

```{r}
(At <- ftrans(A))
```

## Transpose using a R chunk
Converting the inline C++ function to a full C++ function

```{r engine='Rcpp'}
#include <RcppEigen.h>

using namespace Eigen;
using namespace Rcpp;

// [[Rcpp::depends(RcppEigen)]]
// [[Rcpp::export]]
MatrixXi transCpp(const MatrixXi& A) {
  MatrixXi At = A.transpose();
  return At;
}

```


```{r}
require(RcppEigen)

B <- matrix(1:6, ncol = 2)
B
transCpp(B)
```

```{r}
M <- matrix(c(1,2,3,4,5,6,7,8,9), ncol=3)
M
transCpp(M)
```






## XtSx
\[
\hat{\boldsymbol\beta} = \left( \mathbf{X}^\prime \boldsymbol\Sigma^{-1} \mathbf{X} \right)^{-1} \mathbf{X}^\prime \boldsymbol\Sigma^{-1} \mathbf{y}
\]


```{r engine='Rcpp'}
#include <RcppEigen.h>
#include <Eigen/Dense>
#include <Eigen/Cholesky>

using namespace Eigen;     // if we add the namespace, then the class is not needed
using namespace Rcpp;

// [[Rcpp::depends(RcppEigen)]]
// [[Rcpp::export]]
MatrixXd xtSxx(const MatrixXd& X, const MatrixXd& Sigma, const VectorXd& Y) {
  MatrixXd SX = Sigma.llt().solve(X); // A\b by Cholesky's decomposition
  MatrixXd XSX = (X.adjoint())*SX;    // ldlt() for p.s.d. matrices
  VectorXd XSY = (SX.adjoint())*Y;    // llt() for p.d. matrices
  VectorXd beta = XSX.ldlt().solve(XSY);
  return (beta);
}
```

A test case to show the results are the same:

```{r}
require(RcppEigen)

S <- matrix(c(1,.5,.5,1),ncol=2)
Y <- c(3,2)
X <- matrix(c(1,0,1,1), ncol=2)

solve(t(X)%*%solve(S)%*%X)%*%t(X)%*%solve(S)%*%Y
xtSxx(X,S,Y)
```



## Matrix product and cross-product
from the paper `Figure 2` example.

As mentioned in Section 2.2, the '*' operator is overloaded as matrix multiplication for the various matrix and vector classes in $\textbf{Eigen}$.  The C++ code in Figure 2 produces a list containing both the product and cross-product (in the sense of the R function call `crossproduct(A, B)` evaluating $A \top B$) of its two arguments

```{r engine='Rcpp'}
#include <RcppEigen.h>

using namespace Eigen;
using namespace Rcpp;

typedef MatrixXi MapMati;         // declare a type MapMati

// [[Rcpp::depends(RcppEigen)]]
// [[Rcpp::export]]
// prodCpp function is of type List
List prodCpp(const MapMati& B, const MapMati& C) {
  return List::create(Named("B %*% C")        = B * C,
                      Named("crossprod(B, C") = B.adjoint() * C);
}

```

```{r}
B <- matrix(1:4, ncol = 2)
C <- matrix(6:1, nrow = 2)

str(fp <- prodCpp(B, C))

```
> Note  that  the create class  method  for Rcpp::List implicitly  applies Rcpp::wrap
to  its arguments.


## Crossproduct of a single matrix
As shown in the last example, the R function `crossprod` calculates the product of the transpose of its  first argument with its second argument. The single argument form, `crossprod(X)`, evaluates $X^\top X$ .  One could, of course, calculate this product as 

    t(X) %*% X
    
but `crossprod(X)` is roughly twice as fast because the result is known to be symmetric and only one triangle needs to be calculated. The function `tcrossprod` evaluates crossprod(t(X)) without actually forming the transpose.

To express these calculations in Eigen, a SelfAdjointView -which is a dense matrix of which only one triangle is used,  the other triangle being inferred from the symmetry|is created. (The characterization "self-adjoint" is equivalent to symmetric for non-complex matrices).

```{r engine='Rcpp'}
#include <RcppEigen.h>

using namespace Eigen;
using namespace Rcpp;

using Eigen::Lower;

// [[Rcpp::depends(RcppEigen)]]
// [[Rcpp::export]]

List crossprodCpp(const MatrixXi& A) {
  const int m(A.rows()), n(A.cols());
  
  MatrixXi AtA = MatrixXi(n,n).setZero().selfadjointView<Lower>().rankUpdate(A.adjoint());
  MatrixXi AAt = MatrixXi(m, m).setZero().selfadjointView<Lower>().rankUpdate(A);
  
  return List::create(Named("crosprod(A)")  = AtA,
                      Named("tcrossprod(A") = AAt);
}

```

The Eigen class name is __SelfaAjointView__. The method for general matrices that produces such  a  view  is  called `selfadjointView`. Both  require  specification  of  either  the __Lower__ or __Upper__ triangle. For  triangular  matrices  the  class  is __TriangularView__ and  the  method  is `triangularView`. The  triangle  can  be  specified  as _Lower, UnitLower, StrictlyLower, Upper, UnitUpper or StrictlyUpper_. For self-adjoint views the rankUpdate method adds a scalar multiple of
$AA^\top$ to the current symmetric matrix.  The scalar multiple defaults to 1.  The code in Figure 3 produces both $A^\top A$ and $AA^\top$ from a matrix $A$.

```{r}
A <- matrix(1:6, ncol = 2)
crossprodCpp(A)
```

Using help from https://github.com/RcppCore/RcppEigen/blob/master/src/fastLm.cpp



To some, the expressions in Figure 3 to construct  __AtA__  and  __AAt__  are compact and elegant.  To  others they are hopelessly confusing.  If you  nd yourself in the latter group, you just need  to read the expression left to right.  So, for example, we construct  __AAt__  by creating a general  integer  matrix  of  size  $m \times m$  (where  $A$  is  $m  \times n$  ),  ensuring  that  all  its  elements  are  zero,  regarding it as a self-adjoint (i.e., symmetric) matrix using the elements in the lower triangle,  then adding  $AA^\top$  to it and converting back to a general matrix form (i.e., the strict lower  triangle is copied into the strict upper triangle).


In more detail: 

1. `MatrixXi(n, n)` creates an $n \times n$ integer matrix with arbitrary contents


2. `.setZero()`  zeros out the contents of the matrix.

3.  `.selfAdjointView<Lower>()`  causes what follows to treat the matrix as a symmetric  matrix in which only the lower triangle is used, the strict upper triangle being inferred  by symmetry  

4.  `.rankUpdate(A)`  forms the sum  $B  +  AA^\top$  where  $B$  is the symmetric matrix of zeros  created in the previous steps.  

The assignment of this symmetric matrix to the (general)  `MatrixXi`  object  `AAt`  causes the  result to be symmetrized during the assignment.  

For these products one could define the symmetric matrix from either the lower triangle or  the upper triangle as the result will be symmetrized before it is returned.  

To cut down on repetition of  using  statements we gather them in a character variable,  __incl__,  that will be given as the __includes__  argument in the calls to  `cxxfunction`.  We also define a  utility function,  $AtA$, that returns the crossproduct matrix as shown in Figure 4.


## Cholesky decomposition of the crossprod

```{r engine='Rcpp'}

#include <RcppEigen.h>

using namespace Eigen;
using namespace Rcpp;

using Eigen::Lower;

typedef MatrixXd MapMatd;

// [[Rcpp::depends(RcppEigen)]]
// [[Rcpp::export]]

List cholCpp(const MapMatd & A) {
  const MapMatd AtA = A;
  const LLT<MatrixXd> llt(A);

  return List::create(Named("L)")  = MatrixXd(llt.matrixL()),
                      Named("R")   = MatrixXd(llt.matrixU()));
}

```




```{r}
# A <- matrix(1:4, ncol = 2)
# (mm <- Matrix(toeplitz(c(10, 0, 1, 0, 3)), sparse = TRUE)) # 5 x 5
( m <- matrix(c(14,32,7,7),2,2) )

(ll <- cholCpp(m))
```

```{r}
( cm <- chol(m) )
```

```{r}
library(Matrix)
Cholesky(m)
```

