---
title: "Using 'Rcpp' to integrate R and C++"
author: "Guilherme Ludwig"
date: "10/09/2014"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE, error=TRUE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA, error = TRUE)
```

```{r knitr_options, echo=FALSE}
library(knitr)
opts_chunk$set(fig.align='center', fig.path='Rcpp/', warning=FALSE, message=FALSE, echo=TRUE)
```

Source of this document available [here](Rcpp.Rmd).

```{r}
library(Rcpp)
```


First I am going to run an example from Dirk Eddelbuettel's *Seamless R and C++ Integration with Rcpp* (2013); he wrote code to calculate the n-th term of the Fibonacci sequence, a recursive function given by $$F(0) = 0, F(1) = 1, F(n) = F(n-1) + F(n-2)$$

```{r}
fibR <- function(n){
  ## I commented this out because I am not handling errors in the C++ code,
  ## so the code is comparable
  # if (n < 0) {
  #  stop("Argument n must be non-negative integer.\n")
  # } else 
  if (n == 0) {
    return(0)
  } else if (n == 1) {
    return(1)
  } else {
    Recall(n-1) + Recall(n-2)
  }
}
n <- 10:26
time.to.calculate <- sapply(n, function(x) system.time(fibR(x))[3])
plot(n, time.to.calculate, xlab="n", ylab="seconds", 
     main="Fibonacci F calculation time", type="b")
```

This is the C++ code to calculate the same function (I wrote the whole program, but we only need the `fibCpp()` chunk):

```{r eval=FALSE, engine='Rcpp'}
#include <cstdio>

// This is the function we will later import to R
int fibCpp(const int n) {
  if(n == 0){
    return(0);
  } else if(n == 1){
    return(1);
  } else {
    return(fibCpp(n-1) + fibCpp(n-2));
  }
}

/* Main function, required for C++
Just a small script that asks for user input on n
and evaluates/prints it.
*/ 
int main(void){
  int n; // Variables have to be initialized
  int z;
  printf("Write an integer greater or equal than 0:\n");
  scanf("%d", &n); // Assigns an user inputted digit to n
                   // Here %d is a formatted digit
                   // abd &n is the address in the memory for n
  z = fibCpp(n); // Call the function we created previously
  printf("%d\n", z); 
}
```

This code compiles (in Unix shell) by running `g++ filename.cpp -o progname`, and you can execute it by running `./progname`. `filename` and `progname` can be anything, really. But we don't need that for calling C++ code from R. I named mine `fibMain.cpp` and it can be downloaded [here](fibMain.cpp). For just the `fibCpp()` function, it can be downloaded [here](fib.cpp).

Regarding C++ syntax: [this website](http://www.physics.drexel.edu/courses/Comp_Phys/General/C_basics/c_tutorial.html) has a cool introduction to C (not C++). I'm still on the lookout for a nice, 101 level page for C++ (I'm sure many exist). Eddebuettel's book has a cool Appendix on it.

## Using basic Rcpp

**Skip this section and go check "Using Attributes" if you want to start using C++ code immediately. I've only written this because I was coding this markdown file as I read the book, and this was presented first. It is interesting but not necessary anymore**

The C++ code needs an wrapper function. From the [Writing R Extensions](http://cran.r-project.org/doc/manuals/R-exts.html#Interface-functions-_002eCall-and-_002eExternal) manual, and Eddenbuettel's book, one suitable wrapper is

```{r eval=FALSE, engine='Rcpp'}
extern "C" SEXP fibWrapper(SEXP ns) {
  int n = Rcpp::as<int>(ns);
  int fib = fibCpp(n);
  return(Rcpp::wrap(fib));
}
```

So what's going there is that the wrapper function is being declared to return a pointer to an S expression (`SEXP`) and taking an `SEXP` argument `ns`. Rcpp has a function `as<int>` that behaves like `as.integer()` from R, but turns the `ns` object into an integer variable in C++. It evaluates the `fibCpp()` function and `wraps()` the returned integer variable into an object that R understands. Most wrappers should be somewhat like this.

Now the following steps require a compiler; you can guess that I'm using `g++` but `gcc` is also used in the process. I don't know much about compilers beyond basic usage. If you're coding in Linux, you're good (make sure you have the developer tools by running `sudo apt-get install r-base-dev`). If you're on a Mac or, even worse, Windows, apparently the [Rtools](http://cran.at.r-project.org/bin/windows/Rtools/) program has both `g++` and `gcc` available. ~~I did not test it~~, but I used Rtools when teaching [how to create a package](http://pages.stat.wisc.edu/~gvludwig/327-5/package.html) in my Spring '14 class at UW. Update: tested it, it works!

First, I include the `fibWrapper()` function in a different file along with the `fibCpp()` C++ function. This new file is not necessary (the wrapper can be in [fib.cpp](fib.cpp)), but I did it for illustration. I also need to explicitly call `#include <Rcpp.h>`, `using namespace Rcpp;` at the beginning of the file. Everything is in [fibWrap.cpp](fibWrap.cpp).. So, now I'd type in the bash prompt

```{r eval=FALSE}
PKG_CXXFLAGS=`Rscript -e 'Rcpp:::CxxFlags()'` \
PKG_LIBS=`Rscript -e 'Rcpp:::LdFlags()'`  \
R CMD SHLIB fibWrap.cpp
```

and then, after starting up R

```{r}
library(Rcpp)
dyn.load("fibWrap.so")
.Call("fibWrapper", 5) # Evaluates fibWrapper() at 5
.Call("fibWrapper", 9) 
.Call("fibWrapper", 10) 
```

This approach does not work on Windows as is (`.so` is a Linux shared library, you need to create a `.dll` in Windows), but see the next session "Using Attributes". Also these two steps, and the wrapper function, are quite complicated. In fact, Dirk Eddelbuettel is discouraging it. There's a much easier method available, which uses the `Attributes` in the `Rcpp` package.

## Using Attributes

Open my [fibBest.cpp](fibBest.cpp) version; it doesn't have an wrapper anymore, but it includes the `// [[Rcpp::export]]` just before the actual function. If you skipped the last section, go back for a second and read the paragraph about requiring a compiler (both `g++` and `gcc`, install `Rtools` if you're in Windows; might not need if using Rstudio). Anyway, when using Attributes I just need to boot R and run

```{r}
library(Rcpp)
sourceCpp("src/fibBest.cpp")
```

and that makes the function `fibCpp()` available in R!

## Comparing results

This is magic, and speaks for itself. I augmented the values called to show that the function behavior is the same (in terms of complexity) but C++ just does the computation much faster.

```{r}
n2 <- 10:(26+12) # Some extra 12 values
time.to.calculateCpp <- sapply(n2, function(x) system.time(fibCpp(x))[3])

plot(n2, c(time.to.calculate, rep(NA,12)), xlab="n", ylab="seconds", 
     main="Fibonacci F calculation time", type="b")
lines(n2, time.to.calculateCpp, type="b", col="Red")
legend("topleft", inset=0.01, legend=c("R","C++"), pch=1, col=1:2)
```

## NumericVector types and other Rcpp tools

It's somewhat complicated (for someone with a predominantly R-background) to work with vectors and memory allocation in C/C++. I never really understood how to use `malloc()` in C and, while C++ might have some better pointer-based options, I don't know them because I don't know much C++ to begin with. So this example uses the `NumericVector` type from Rcpp to reproduce a function similar to R's `dnorm()`, with only some basic arguments (namely, `x`, `mean` and `sd` in R).

```{r eval=FALSE}
#include <Rcpp.h>
#include <cmath>        // std::exp(double)
#include <valarray>     // std::valarray, std::exp(valarray)

using namespace Rcpp;

// [[Rcpp::export]]
NumericVector dnormCpp(NumericVector x, double mu, double sigma) {
   // const double MYPI = 3.141592;
   // I think PI is available in cmath
   int n = x.size();
   NumericVector d(n);
   NumericVector ret(n);
   for(int i = 0; i < n; i++){
     d[i] = (x[i] - mu)/sigma;
     d[i] *= d[i];
     ret[i] = exp(-0.5*d[i])/(sqrt(2*PI)*sigma);
   }
   return(ret);
}
```

Download the code [here](dnormCpp.cpp). It's not much faster than the regular R version (which, to be fair, includes a lot of error checking and is probably faster at its core anyway, since it's written in C), but this example is a good start for functions that involve the Normal density and some loops.

Note also that I'm not calculating values at the tails of the density or anything, which require smarter algorithms for better accuracy, and should be available in `dnorm()` already.

```{r}
library(Rcpp)
sourceCpp("src/dnormCpp.cpp")

all.equal(dnorm(1:10, 5, 1), dnormCpp(1:10, 5,1))

n <- 20
timeCpp <- timeR <- numeric(n)
for(k in 1:n){
  timeCpp[k] <- system.time({dnormCpp(1:(10000*n), 10000*n/2, 10000*n)})[3]
  timeR[k] <- system.time({dnorm(1:(10000*n), 10000*n/2, 10000*n)})[3]
}

plot(10000*(1:n), timeR, xlab="n", ylab="seconds", ylim=c(0, max(timeR)),
     main="dnorm() v. dnormCpp() times", type="b")
lines(10000*(1:n), timeCpp, type="b", col="Green")
legend("bottomleft", inset=0.01, legend=c("R","C++"), pch=1, col=c("Black", "Green"))
```

## More complicated example using RcppEigen

Basically I am going to implement a fast way to calculate

\[
\hat{\boldsymbol\beta} = \left( \mathbf{X}^\prime \boldsymbol\Sigma^{-1} \mathbf{X} \right)^{-1} \mathbf{X}^\prime \boldsymbol\Sigma^{-1} \mathbf{y}
\]

using the `RcppEigen` package. Eigen is a super fast linear algebra library for C++, that Jared Huling told me about. There's an useful [quick reference](http://eigen.tuxfamily.org/dox/AsciiQuickReference.txt) on Eigen, as well as a in-depth [tutorial](http://eigen.tuxfamily.org/dox/group__TutorialMatrixClass.html). The quick reference is very conveniently linking commands to MATLAB scripts. :)

First, we need [Eigen](http://eigen.tuxfamily.org/index.php?title=Main_Page) on the system. It can be downloaded [here](http://eigen.tuxfamily.org/index.php?title=Main_Page#Download). I think it is possible to `sudo apt-get install libeigen3-dev` instead, but it didn't work for me. Instead I ran the following:

```{r eval=FALSE, engine='Rcpp'}
cd ~/Downloads
tar xzf eigen-eigen-1306d75b4a21.tar.gz
cd eigen-eigen-1306d75b4a21
sudo cp -r Eigen /usr/include/
```

Here's my implementation of the calculation, using Rcpp-Eigen. I saved the source in [xtSx.cpp](xtSx.cpp):

```{r eval=FALSE, engine='Rcpp'}
#include <RcppEigen.h>
#include <Eigen/Dense>
#include <Eigen/Cholesky>

using namespace Eigen;
using namespace Rcpp;

// [[Rcpp::depends(RcppEigen)]]
// [[Rcpp::export]]
MatrixXd xtSx(const MatrixXd& X, const MatrixXd& Sigma, const VectorXd& Y) {
  MatrixXd SX = Sigma.llt().solve(X); // A\b by Cholesky's decomposition
  MatrixXd XSX = (X.adjoint())*SX;    // ldlt() for p.s.d. matrices
  VectorXd XSY = (SX.adjoint())*Y;    // llt() for p.d. matrices
  VectorXd beta = XSX.ldlt().solve(XSY);
  return (beta);
}
```

A test case to show the results are the same:

```{r}
require(RcppEigen)

sourceCpp("src/xtSx.cpp")
S <- matrix(c(1,.5,.5,1),ncol=2)
Y <- c(3,2)
X <- matrix(c(1,0,1,1), ncol=2)
solve(t(X)%*%solve(S)%*%X)%*%t(X)%*%solve(S)%*%Y
xtSx(X,S,Y)
```

Keeping track of time, using two versions of R implementation (the straightforward, above, and a clever one using `crossprod`)

```{r}
xtSxR <- function(X, S, Y){
  return(solve(t(X)%*%solve(S)%*%X)%*%t(X)%*%solve(S)%*%Y)
}
xtSxRclever <- function(X, S, Y){
  return(solve(crossprod(X, solve(S, X)), crossprod(X, solve(S, Y))))
}
```

And a simulation for a large value of $p = 10$, $n > p$

```{r}
p <- 10
n <- seq(200, 500, by=20)
timeCpp <- numeric(length(n))
timeR <- numeric(length(n))
timeRclever <- numeric(length(n))
k <- 0
set.seed(1)
for(i in n){
  k <- k + 1
  Y <- rnorm(i)
  X <- matrix(rnorm(i*p), ncol=p)
  S <- matrix(0, i, i)
  for(j in 1:i){
    S[j,] = 0.5^abs(j - 1:i)
  }
  timeCpp[k] <- system.time(xtSx(X,S,Y))[3]
  timeR[k] <- system.time(xtSxR(X,S,Y))[3]
  timeRclever[k] <- system.time(xtSxRclever(X,S,Y))[3]
}
```

This graph shows how much better the C++ version is:

```{r}
plot(n, timeR, xlab="n", ylab="seconds", ylim=c(0, max(timeR)),
     main="Generalized LS calculation time", type="b")
lines(n, timeRclever, type="b", col="Red")
lines(n, timeCpp, type="b", col="Blue")
legend("topleft", inset=0.01, legend=c("R","Rclever","C++"), pch=1, col=c("Black", "Red", "Blue"))
```

Relative difference:

```{r}
all.equal(as.numeric(xtSx(X,S,Y)), 
          as.numeric(xtSxR(X,S,Y)), 
          as.numeric(xtSxRclever(X,S,Y)))
```

At $p=10, n=500$, the improvement in speed is about `r round(max(timeR)/max(timeCpp))`-fold! It can be made faster, though. I am going to try modify my example inspired in the $\mathbf{X}^\prime \mathbf{X}$ example from Bates and Eddelbuettel's "Fast and Elegant Numerical Linear Algebra Using the RcppEigen Package" (<i>Journal of Statistical Software</i>, 2013) and on some suggestions that [Jared](https://github.com/jaredhuling) shared with me. Code is [here](xtSxMod.cpp).

```{r}
sourceCpp("src/xtSxMod.cpp")

timeCppFast <- numeric(length(n))
k <- 0
set.seed(1)
for(i in n){
  k <- k + 1
  Y <- rnorm(i)
  X <- matrix(rnorm(i*p), ncol=p)
  S <- matrix(0, i, i)
  for(j in 1:i){
    S[j,] = 0.5^abs(j - 1:i)
  }
  timeCppFast[k] <- system.time(xtSxMod(X,S,Y))[3]
}
```

Now, the graph shows that this modified version is not faster. I think that it's because I'm calling the solver method three times, as opposed to two before, so the `selfAdjointView` method is not really helping. I should learn how to profile C++ code.

```{r}
plot(n, timeR, xlab="n", ylab="seconds", ylim=c(0, max(timeR)),
     main="Generalized LS calculation time", type="b")
lines(n, timeRclever, type="b", col="Red")
lines(n, timeCpp, type="b", col="Blue")
lines(n, timeCppFast, type="b", col="Orange")
legend("topleft", inset=0.01, legend=c("R","R Fast","C++","C++ Fast"), pch=1, col=c("Black", "Red", "Blue", "Orange"))
```