---
title: "software-inspection"
output: html_notebook
---

Loading the Matlab software array:
```{r}
software.mat <- readMat("./data/software.mat", fixNames = TRUE)
```


```{r}
names(software.mat)
```

The file software.mat contains the preparation time in minutes (prepage, prepsloc), the total work hours in minutes for the meeting (mtgsloc), and the number of defects found (defpage, defsloc). 

The variables are normalized by the size of the inspection (the number of pages or SLOC – single lines of code).
 
```{r}
prepsloc <- software.mat$prepsloc
defsloc <- software.mat$defsloc
```


```{r, fig.asp=1}
plot(prepsloc, defsloc)
```

## Transformation

the relationship between the variables is
difficult to understand. 

```{r, fig.asp=1}
# first data transformation
X = log(prepsloc)
Y = log(defsloc)

# plot the transformed data
plot(X, Y, 
     xlab = "Log PrepTime/SLOC",
     ylab = "Log Defects/SLOC"
)

```

We now have a better idea of the relationship between these two variables.

"Some transformations of the data may lead to insights or discovery of
structures that we might not otherwise see. However, as with any analysis,
we should be careful about creating something that is not really there, but is
just an artifact of the processing. Thus, in any application of EDA, the analyst
should go back to the subject area and consult domain experts to verify and
help interpret the results."

"


