---
title: "projection. Example 2.1"
output:
  html_document: default
  html_notebook: default
---

# Objective
In this example we want to show the dimension reduction of the data. We start with a 2-D data set and will reduce it to a 1-D data set by projecting the x,y coordinates of the data set over the x-axis.


We will be projecting our data set onto a line that is $\theta$ radians from the horizontal.

$\theta$ in the example is $\frac{\pi}{3}$.

The projection matrix is given by:

$p = \begin{pmatrix}
(cos \theta)^2 & cos \theta sin \theta \\ 
 cos \theta sin \theta & (sin \theta)^2
 \end{pmatrix}$

```{r}
# enter data as rows
X = matrix(c(4, 3, -4, 5), nrow=2, byrow = TRUE)

# get theta
theta <- pi /3

# obtain the projection matrix
c2 <- cos(theta)^2
cs <- cos(theta)*sin(theta)
s2 <- sin(theta)^2
P <- array(c(c2, cs, cs, s2), dim = c(2,2))
P
```

```{r, fig.asp=0.8}
source("../matlab.R")

Xp <- X %*% P

plot(Xp[, 1], Xp[, 2], 
     xlim = c(-8, 8),
     ylim = c(-8, 8)
     )

Px <- matrix(c(1, 0), nrow=2)
Xpx <- X %*% Px

projectionMatrixLine(P, col = 'green')   # draw the projection line at angle

abline(h=0, v=0)          # lines at (0, 0)
points(Xpx[1], 0, pch=8)  # plot point projections
points(Xpx[2], 0, pch=8)
```

We can see now the ortogonal projections on the x-axis.

```{r}
X
```

