---
title: "Example 2.4 - Factor Analysis"
output:
  pdf_document: 
    latex_engine: xelatex
  html_notebook: default
  html_document: default
---

```{r setup, include=FALSE, error=TRUE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA, error = TRUE)
```

Source code: http://apt.kde-redhat.org/fedora/SOURCES/matlab/matlab/toolbox/stats/

In this example, we examine some data provided with the Statistics Toolbox,
called `stockreturns`. An alternative analysis of these data is provided in
the Statistics Toolbox User ’s Guide. The data set consists of 100 observations,
representing the percent change in stock prices for 10 companies. Thus, the
data set has n = 100 observations and p = 10 variables. It turns out that the first
four companies can be classified as technology, the next three as financial,
and the last three as retail. We can use factor analysis to see if there is any
structure in the data that supports this grouping. We first load up the data set
and perform factor analysis using the function factoran.

```{r}
library(R.matlab)
source("../matlab.R")

# read the Matlab array
stockreturns.mat <- readMat("../data/stockreturns.mat", fixNames = TRUE)
names(stockreturns.mat)
```

```{r}
stocks <- stockreturns.mat$stocks
```

